\documentclass[../main.tex]{subfiles}

\begin{document}

\noindent Clinical trials serve a critical role in the generation of medical evidence and furthering of biomedical research. In order to identify potential participants, investigators publish eligibility criteria, such as past history of certain conditions, treatments, or laboratory tests. Patients meeting a trial's eligibility criteria are considered potential candidates for recruitment. Recruitment of participants remains, however, a major barrier to successful trial completion \cite{richesson2013electronic}, and manual chart review of hundreds or thousands of patients to determine a candidate pool can be prohibitively labor- and time-intensive. While cohort discovery tools such as Leaf \cite{dobbins2019leaf} or i2b2 \cite{murphy2010serving} can serve to assist in finding participants meeting eligibility criteria, such tools nonetheless often have significant learning curves, and certain complex queries may simply be impossible due to structural limitations on the types of possible queries presented.

An alternative approach which holds promise is the use of natural language processing (NLP) to automatically analyze eligibility criteria and generate queries to find patients in databases. NLP-based approaches have the advantage of obviating potential learning curves of tools such as Leaf, while leveraging existing eligibility criteria composed in a free-text format researchers are already familiar with.

The goal of this project is the development of an application called LeafAI, which if successful will enable the discovery of patients meeting criteria for clinical trials and general purpose biomedical research using natural language and generating queries for virtually any clinical database structure.

\subsection{Aim 1. Creation of Gold Standard Clinical Trials Data Set}
In Aim 1 we seek to create a gold standard human-annotated data set of clinical trial eligibility criteria. Our annotation schema expands on previous work \cite{kury2020chia, kang2017eliie} while capturing significantly greater semantic granularity in order to facilitate accurate query generation and enable handling of complex and non-specific criteria. Predictive models trained on the resulting corpus enable subsequent work in Aim 2.

\subsection{Aim 2. Query Generation Method Development and Evaluation}
A successful application for cohort discovery using NLP for end-users hinges on development of robust, explainable methods for generating database queries. In Aim 2 we focus on creation of a novel database model-agnostic approach for cohort discovery query generation using NLP, including an integrated knowledge base (KB), Sequence to Sequence- (Seq2Seq) based logical transformations of eligibility criteria, normalization, and semantic metadata mapping (SMM) of database structure using concepts within the Unified Medical Language System (UMLS). To evaluate our system, we analyze queries generated for 8 past clinical trials at our institution and compared enrolled patients to those found our systems compared to a human programmer.

\subsection{Aim 3. Web Application Development and Evaluation}
Aim 3 will focus on the development of a web application enabling automatic query generation using user-entered free-text eligibility criteria. The LeafAI web application will utilize a chat-like interface to enable iterative, explainable, interactive query generation. Users will then be able to edit and re-execute queries based on their findings. We will evaluate the effectiveness of the web application by comparing both (1) usability evaluations and (2) performance of generated queries in finding enrolled patients between LeafAI and our previous non-NLP based cohort discovery tool, Leaf.

\newpage

\end{document}