%  ========================================================================
%  Copyright (c) 1985 The University of Washington
%
%  Licensed under the Apache License, Version 2.0 (the "License");
%  you may not use this file except in compliance with the License.
%  You may obtain a copy of the License at
%
%      http://www.apache.org/licenses/LICENSE-2.0
%
%  Unless required by applicable law or agreed to in writing, software
%  distributed under the License is distributed on an "AS IS" BASIS,
%  WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
%  See the License for the specific language governing permissions and
%  limitations under the License.
%  ========================================================================
%

% Documentation for University of Washington thesis LaTeX document class
% by Jim Fox
% fox@washington.edu
%
%    Revised 2020/02/24, added \caption()[]{} option.  No ToC.
%
%    Revised for version 2015/03/03 of uwthesis.cls
%    Revised, 2016/11/22, for cleanup of sample copyright and title pages
%
%    This document is contained in a single file ONLY because
%    I wanted to be able to distribute it easily.  A real thesis ought
%    to be contained on many files (e.g., one for each chapter, at least).
%
%    To help you identify the files and sections in this large file
%    I use the string '==========' to identify new files.
%
%    To help you ignore the unusual things I do with this sample document
%    I try to use the notation
%       
%    % --- sample stuff only -----
%    special stuff for my document, but you don't need it in your thesis
%    % --- end-of-sample-stuff ---


%    Printed in twoside style now that that's allowed
%
 
\documentclass [11pt, proquest] {uwthesis}[2020/02/24]

% --- Nic-added packages ---
\usepackage{subfiles}
\usepackage{url}
\usepackage{graphicx}
\usepackage{float}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{changepage}
\usepackage{caption}
\usepackage{multirow}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{makecell}
\usepackage[math]{cellspace}
\usepackage{tikz}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{color}
\usepackage[ruled, linesnumbered]{algorithm2e}
%\usepackage{refcheck} % use with \nocite{*} to get unused references
% --- end Nic-added packages ---
 
%
% The following line would print the thesis in a postscript font 

% \usepackage{natbib}
% \def\bibpreamble{\protect\addcontentsline{toc}{chapter}{Bibliography}}

\setcounter{tocdepth}{1}  % Print the chapter and sections to the toc
 

% ==========   Local defs and mods
%

% --- sample stuff only -----
% These format the sample code in this document

\usepackage{alltt}  % 
\newenvironment{demo}
  {\begin{alltt}\leftskip3em
     \def\\{\ttfamily\char`\\}%
     \def\{{\ttfamily\char`\{}%
     \def\}{\ttfamily\char`\}}}
  {\end{alltt}}
 
% metafont font.  If logo not available, use the second form
%
% \font\mffont=logosl10 scaled\magstep1
\let\mffont=\sf
% --- end-of-sample-stuff ---
 



\begin{document}
 
\prelimpages
 
\Title{Explainable query generation for cohort discovery \\ and biomedical reasoning using natural language}
\Author{Nicholas J. Dobbins}
\Year{2023}
\Program{Biomedical and Health Informatics}

\Chair{Meliha Yetisgen}{Professor}{Biomedical and Health Informatics}
\Signature{H. Nina Kim, MD, MSc}
\Signature{Trevor Cohen, MBChB, PhD}

\copyrightpage

\titlepage  

\setcounter{page}{-1}

\abstract{
    \subfile{Sections/0_abstract}
}
 
\tableofcontents
\listoffigures
\listoftables  
 
\chapter*{List of Abbreviations}    
\addcontentsline{toc}{chapter}{List of Abbreviations}
\thispagestyle{plain}
\subfile{Sections/0_abbreviations}
 
\acknowledgments{% \vskip2pc
  I want to begin by thanking my advisor, Meliha Yetisgen. At a time when I knew next to nothing about natural language processing, she kindly and patiently met with me and explained the basics and suggested what I should focus on learning. She was among the first to suggest I get a PhD, wrote not one but two recommendations for my graduate school application (first as non-matriculated, then as a formal PhD student), and has provided ongoing constructive feedback, advice, support, and opportunities for much of my career. I think there are few people in life that we can objectively say we have been changed by, but in mine she is certainly one of them. I'm eternally grateful.

  I also want to thank the members of my committee, Nina Kim, Trevor Cohen, and Fei Xia. Nina was among the first physicians I interacted with at UW, and her feedback and suggestions over the years have been extremely helpful. Trevor and Fei are both truly excellent teachers and passionate about their work.

  Sean Mooney and Peter Tarczy-Hornoch have also been critical to my PhD journey. I'll never forget the day that I showed Peter\textemdash who happened to be passing by\textemdash a demo of the prototype of Leaf. That Peter was willing to sit and talk with me about how it worked for over an hour meant the world to me (particularly as a former History major with never-ending imposter syndrome). Sean has been unfailingly supportive and a wonderful mentor.

  Last, I'm incredibly grateful to my wife, Akiko. Through the years of working full-time while pursuing my PhD, she has helped me stay grounded and focused on our family and children, who I am intensely proud of and who will always be my greatest joy and priority. I can only hope that they grow to enjoy the same love of learning and imagining \textit{what could be?} that I think make life interesting and worth living.  
}

\dedication{\begin{center}To my children, Elly, Ami, and Sage.\end{center}}

%
% end of the preliminary pages

\textpages

\chapter{Introduction}
\label{chap:introduction}
\subfile{Sections/1_introduction}

\chapter{Background}
\label{chap:background}
\subfile{Sections/2_background}

\chapter{Leaf Clinical Trials Corpus}
\label{chap:lct_corpus}
\subfile{Sections/3_lct_corpus}

\chapter{Leaf Logical Forms Corpus}
\label{chap:llf_corpus}
\subfile{Sections/4_llf_corpus}

\chapter{Knowledge Base}
\label{chap:kb}
\subfile{Sections/5_knowledge_base}

\chapter{Semantic Metadata Mapping}
\label{chap:smm}
\subfile{Sections/6_smm}

\chapter{Query Generation}
\label{chap:query_generation}
\subfile{Sections/7_query_generation}

\chapter{Web Application}
\label{chap:web_appplication}
\subfile{Sections/8_web_application}

\chapter{Conclusions}
\label{chap:conclusions}
\subfile{Sections/9_conclusions}
 
\printendnotes

%
% ==========   Bibliography
%
%\nocite{*}   % include everything in the uwthesis.bib file
\bibliographystyle{ieeetr}
\bibliography{main}
%
% ==========   Appendices
%
\appendix
\raggedbottom\sloppy
 
% ========== Appendix A
%\chapter{Appendix A}
%\label{sec:appendix_a}
%\subfile{Sections/10_appendix}


\end{document}