%\documentclass[fleqn,10pt]{wlscirep}
\documentclass[sort&compress, numafflabel]{elsarticle}

% SDOH paper packages
\usepackage[]{natbib}
\usepackage[breaklinks,hidelinks]{hyperref}
\usepackage{times}
\usepackage{latexsym}
\renewcommand{\UrlFont}{\ttfamily\small}
\usepackage{geometry}
\usepackage{subfiles}
\usepackage{multirow}
\usepackage{array}
\usepackage{hyphenat}
\usepackage{subcaption}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{enumitem}
\usepackage{url}
\def\UrlBreaks{\do\/\do-\do\_}
\usepackage[color=yellow, textsize=small]{todonotes}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{float}
\usepackage[english]{babel}
\usepackage{blindtext}
\usepackage{textcomp}
\usepackage{soul,color}
\usepackage{pifont}

\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{}%
 \let\@evenfoot\@oddfoot}
\makeatother
\usepackage{epstopdf}
\epstopdfDeclareGraphicsRule{.tif}{png}{.png}{convert #1 \OutputFile}
\AppendGraphicsExtensions{.tif}

\usepackage{titlesec}
\titleformat{\section}
      {\normalfont\bfseries}
      {\thesection}
      {0ex}
      {\MakeUppercase}
      %[\vspace{1ex}]

\titleformat{\subsection}
      {\normalfont\bfseries}
      {\thesection}
      {0ex}
      {}
      %[\vspace{1ex}]

\titleformat{\subsubsection}
      {\normalfont}
      {\thesection}
      {0ex}
      {}

\bibliographystyle{bst/vancouver2}   

\newcommand\blfootnote[1]{%
  \begingroup
  \renewcommand\thefootnote{}\footnote{#1}%
  \addtocounter{footnote}{-1}%
  \endgroup
}

\usepackage{setspace}
\doublespacing
\newif\ifsubfile
\subfiletrue
\newif\iftif
\tiffalse

% prev packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lineno}
\usepackage[numbers]{natbib}
\usepackage{subfiles}
\usepackage{graphicx}
\usepackage{float}
\usepackage{hyperref}
\usepackage{mathtools}
%\usepackage{graphicx}
\usepackage{changepage}
\usepackage{caption}
\usepackage{multirow}
%\usepackage{parskip}
\usepackage{titlesec}
\usepackage{amsmath}

\usepackage{makecell}

\usepackage[math]{cellspace}
\usepackage{tikz}

\usepackage{url}
\def\UrlBreaks{\do\/\do-\do\_}
\usepackage{colortbl}
\linenumbers
\nolinenumbers

\usepackage{listings}
\usepackage{xcolor}

\title{The LeafAI query engine: a data model-agnostic query generator for cohort discovery and multi-hop biomedical reasoning using natural language}

\author[bime,rit]{Nicholas J Dobbins\textsuperscript{+}\blfootnote{\textsuperscript{+}Corresponding author: Nicholas Dobbins, MLIS, Department of Biomedical Informatics and Medical Education, University of Washington, 1851 NE Grant Ln, Seattle, WA 98195, USA; ndobb@uw.edu.}}

\ead{}
\author[ischool]{Bin Han}\ead{}
\author[bime]{Weipeng Zhou}\ead{}
\author[med]{Kristine Lan}\ead{}
\author[med]{H. Nina Kim}\ead{}
\author[med]{Robert Harrington}\ead{}
\author[ist]{Özlem Uzuner}\ead{}
\author[bime]{Meliha Yetisgen}\ead{}

\address[bime]{Department of Biomedical Informatics \& Medical Education, University of Washington, Seattle, WA, USA}
\address[rit]{Department of Research IT, UW Medicine, University of Washington, Seattle, WA, USA}
\address[med]{Department of Medicine, University of Washington, Seattle, WA}
\address[ischool]{Information School, University of Washington, Seattle, WA}
\address[ist]{Department of Information Sciences and Technology, George Mason University, Fairfax, VA \blfootnote{ \\ Word count: 3991} \blfootnote{\\ Keywords: clinical trials, natural language processing, machine learning, electronic health records, cohort definition}}

\begin{document}
\subfilefalse

\newpageafter{author}

\begin{abstract}
    \subfile{sections/abstract}
\end{abstract}

\begin{keyword}
clinical trials, natural language processing, machine learning, electronic health records, cohort definition
\end{keyword}

\maketitle

\pagebreak

\section*{Introduction}
\label{sec:background}
\subfile{sections/introduction}

\section*{Background and Significance}
\label{sec:background}
\subfile{sections/background}

\section*{Materials and Methods}
\label{sec:methods}
\subfile{sections/methods}

\section*{Results}
\label{sec:results}
\subfile{sections/results}

\section*{Discussion}
\label{sec:discussion}
\subfile{sections/discussion}

\section*{Conclusion}
\label{sec:conclusion}
\subfile{sections/conclusion}

%TC:ignore 
\section*{Acknowledgements} 

This study was supported in part by the National Library of Medicine under Award Number R15LM013209 and by the National Center for Advancing Translational Sciences of National Institutes of Health under Award Number UL1TR002319. Experiments were run on computational resources generously provided by the UW Department of Radiology. The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health. 

\section*{Author contributions statement}

ND is the developer of LeafAI and wrote the majority of the manuscript. BH and WZ annotated the LLF dataset and contributed to the annotation schema. KL served as the human database programmer, and NK and RH advised on study design. OU and MY advised on strategies for query generation and NLP architectures. All authors contributed to the interpretation of the data, manuscript revisions, and intellectual value to the manuscript.

\section*{Competing interests}

The authors declare no competing interests.

\bibliography{sample}

\section*{Appendix}
\label{sec:appendix}
\subfile{sections/appendix}

%TC:endignore 

\end{document}